Modeling Lake Trophic State: A Data Mining Approach

### 2014-09-15 inst/doc/manuscript.Rmd
devtools rebuild with new manuscript.latex template from manuscriptPackage

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
authors: Jeffrey W. Hollister; W. Bryan Milstead; Betty J. Kreakie
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: article
spacing: doublespacing
linenumbers: no
bibliography: components/LakeTrophicModelling.bib
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

*US Environmental Protection Agency, Office of Research and Development, National Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02879, USA*

*corresponding author:hollister.jeff@epa.gov*

\begin{abstract}
Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to paramterize the model.  This limits the application of these models to lakes with existing and, more importanly, available water quality data.  To expand our ability to predict in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database, land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake tropic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two models of trophic state (as determined by chlorophyll *a* concentration).   The first model estimates trophic state with *in situ* as well as universally available data and the second model uses universally available data only. For each of these models we used three separate trophic state categories, for a total of six models. Overall accuracy for the *in situ* and universal data models ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state.  For the universal data only models, Overall accuraccy ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it is believed that the presence and abundance of cyanobacteria is strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria biovolume and the measured and predicted trophic state.  Expanding these preliminary results to include cyanobacteria taxa indicates that cyanobacteria are significantly more likely to be found in highly eutrophic lakes. These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa. 
\end{abstract}

@@

```{r nlaMap, echo=FALSE, fig.cap="Map of the distribution of National Lakes Assesment Sampling locations\\label{fig:nlaMap}"}
state<-map_data('state')
lakes_alb<-data.frame(hkm2014Data[["AlbersX"]],hkm2014Data[["AlbersY"]])
p4s<-"+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
ll<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
lakes_alb_sp<-SpatialPoints(coordinates(lakes_alb),proj4string=CRS(p4s))
lakes_dd<-spTransform(lakes_alb_sp,CRS=CRS(ll))
lakes_dd<-data.frame(coordinates(lakes_dd))
mycolor<-wes.palette(5,"Zissou2")
mycolor<-c(mycolor[1],mycolor[2],mycolor[4])
names(lakes_dd)<-c("long","lat",c())
nlaMap(state,lakes_dd,mycolor)
```

@@

```{r trophicStateTable, results='asis', echo=FALSE, tab.cap="Chlorophyll a based trophic state cut-offs\\label{tab:trophicStateTable}"}
ts_4<-c("oligo","meso","eu","hyper")
ts_3<-c("oligo","meso/eu","meso/eu","hyper")
ts_2<-c("oligo/meso","oligo/meso","eu/hyper","eu/hyper")
co<-c("<= 0.2",">2-7",">7-30",">30")
xdf<-data.frame(ts_4,ts_3,ts_2,co)
names(xdf)<-c("Trophic State (4)","Trophic State (3)","Trophic State (2)","Cut-off")
kable(xdf)
```

@@

```{r VarSel_Model1,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 1\\label{tab:VarSel_Model1}"}
kable(sumTable(all_ts4[1:100]),row.names=F)
```

```{r Confusion_Model1,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 1\\label{tab:Confusion_Model1}"}
kable(formatC(round(all_ts4_rf$confusion,2)),row.names=F)
```

@@

```{r Importance_Model1,results="asis",echo=FALSE, fig.cap="Importance plot for Model 1\\label{fig:Importance_Model1}"}
all_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts4_vars]
importancePlot(all_ts4_rf,'gini',size=10,aes(colour=all_ts4_color))
```

@@

```{r VarSel_Model2,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 2"}
kable(sumTable(all_ts3[1:100]),row.names=F)
```

```{r Confusion_Model2,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 2"}
kable(formatC(round(all_ts3_rf$confusion,2)),row.names=F)
```

```{r Importance_Model2,results="asis",echo=FALSE, fig.cap="Importance plot for Model 2"}
all_ts3_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts3_vars]
importancePlot(all_ts3_rf,'gini',size=10,aes(colour=all_ts3_color))
```

@@

```{r all_TS2_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts2_vars<-unique(unlist(all_ts2[1:100]))
all_ts2_rf<-runRandomForest(hkm2014Data[all_ts2_vars],hkm2014Data$TS_CHLA_2, ntree=10000,importance=TRUE,proximity=TRUE)
```

@@

```{r VarSel_Model3,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 3"}
kable(sumTable(all_ts2[1:100]),row.names=F)
```

```{r Confusion_Model3,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 3"}
kable(formatC(round(all_ts2_rf$confusion,2)),row.names=F)
```

```{r Importance_Model3,results="asis",echo=FALSE, fig.cap="Importance plot for Model 3"}
all_ts2_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts2_vars]
importancePlot(all_ts2_rf,'gini',size=10,aes(colour=all_ts2_color))
```

@@

```{r VarSel_Model4,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 4" }
kable(sumTable(gis_ts4[1:100]),row.names=F)
```

```{r Confusion_Model4,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 4" }
kable(formatC(round(gis_ts4_rf$confusion,2)),row.names=F)
```

```{r Importance_Model4,results="asis",echo=FALSE, tab.cap="Importance plot for Model 4"}
gis_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts4_vars]
importancePlot(gis_ts4_rf,'gini',size=10,aes(colour=gis_ts4_color))
```

@@

```{r VarSel_Model5,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 5"}
kable(sumTable(gis_ts3[1:100]),row.names=F)
```

```{r Conufsion_Model5,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 5"}
kable(formatC(round(gis_ts3_rf$confusion,2)),row.names=F)
```

```{r Importance_Model5,results="asis",echo=FALSE, fig.cap="Importance plot for Model 5"}
gis_ts3_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts3_vars]
importancePlot(gis_ts3_rf,'gini',size=10,aes(colour=gis_ts3_color))
```

@@

```{r VarSel_Model6,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 6"}
kable(sumTable(gis_ts2[1:100]),row.names=F)
```

```{r Confusion_Model6,results="asis",echo=FALSE, tab.cap="Random forest confusion matrix for Model 6"}
kable(formatC(round(gis_ts2_rf$confusion,2)),row.names=F)
```

```{r Importance_Model6,results="asis",echo=FALSE, fig.cap="Importance plot for Model 6"}
gis_ts2_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts2_vars]
importancePlot(gis_ts2_rf,'gini',size=10,aes(colour=gis_ts2_color))
```

@@

```{r ts_4_biov, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 4 trophic state classes"}
plotCdf(hkm2014Data$TS_CHLA_4,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:9],
        y='Percent',x='Log10(Cyanobacteria Biovolume)',
        title=expression(paste('CDF for Chlorophyll ', 
                                italic("a"),' Trophic States (4 Categories)')),
        color="Trophic State\nCategories")
```
\newpage
```{r ts_3_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 3 trophic state classes" }
plotCdf(hkm2014Data$TS_CHLA_3,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:8],
       y='Percent',x='Log10(Cyanobacteria Biovolume)', 
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (3 Categories)')),
       color="Trophic State\nCategories")
```
\newpage
```{r ts_2_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 2 trophic state classes"}
plotCdf(hkm2014Data$TS_CHLA_2,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:7],
       y='Percent',x='Log10(Cyanobacteria Biovolume)',
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (2 Categories)')),
       color="Trophic State\nCategories")
```
\newpage
```{r scatterplot,echo=FALSE, fig.width=7, fig.cap="Cholorphyll *a* and cyanobacteria biovolume scatterplot"}
scp_df<-data.frame(chla=hkm2014Data[["CHLA"]],biovp1=hkm2014Data[["sumBioV"]]+1)
scatterPlot(scp_df,xvar="chla",yvar="biovp1",zissou[8],zissou[7],zissou[6],
             title=expression(paste("Chlorophyll ", 
                                    italic("a")," and Cyanobacteria Relationship")),
             x=expression(paste('Log10(Chl ', italic("a"),')')),
             y="Log10(Cyanobaterial Biovolumes + 1)")
```

>>>>

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author: 
- name: Jeffrey W. Hollister
  affilnum: 1
  email: hollister.jeff@epa.gov
- name: W. Bryan Milstead
  affilnum: 1
- name: Betty J. Kreakie
  affilnum: 1
affiliation:
- affilnum: 1
  affil: US Environmental Protection Agency, Office of Research and Development, National     Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02882, USA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: article
spacing: doublespacing
linenumbers: no
bibliography: components/LakeTrophicModelling.bib

---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

```{r analysis , include=FALSE, echo=FALSE}
#All analysis in here, that way all bits of the paper have access to the final objects
#Place tables and figures and numerical results where they need to go.
```

\begin{singlespace}
\begin{abstract}
Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to paramterize the model.  This limits the application of these models to lakes with existing and, more importanly, available water quality data.  To expand our ability to predict in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database, land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake tropic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two models of trophic state (as determined by chlorophyll *a* concentration).   The first model estimates trophic state with *in situ* as well as universally available data and the second model uses universally available data only. For each of these models we used three separate trophic state categories, for a total of six models. Overall accuracy for the *in situ* and universal data models ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state.  For the universal data only models, Overall accuraccy ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it is believed that the presence and abundance of cyanobacteria is strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria biovolume and the measured and predicted trophic state.  Expanding these preliminary results to include cyanobacteria taxa indicates that cyanobacteria are significantly more likely to be found in highly eutrophic lakes. These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa. 
\end{abstract}
\end{singlespace}

@@

```{r nlaMap, echo=FALSE, fig.cap="Map of the distribution of National Lakes Assesment Sampling locations\\label{fig:nlaMap}",cache=TRUE}
state<-map_data('state')
lakes_alb<-data.frame(hkm2014Data[["AlbersX"]],hkm2014Data[["AlbersY"]])
p4s<-"+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
ll<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
lakes_alb_sp<-SpatialPoints(coordinates(lakes_alb),proj4string=CRS(p4s))
lakes_dd<-spTransform(lakes_alb_sp,CRS=CRS(ll))
lakes_dd<-data.frame(coordinates(lakes_dd))
mycolor<-wes.palette(5,"Zissou2")
mycolor<-c(mycolor[1],mycolor[2],mycolor[4])
names(lakes_dd)<-c("long","lat",c())
nlaMap(state,lakes_dd,mycolor)
```

@@

```{r trophicStateTable, results='asis', echo=FALSE, tab.cap="Chlorophyll a based trophic state cut-offs\\label{tab:trophicStateTable}",cache=TRUE}
ts_4<-c("oligo","meso","eu","hyper")
ts_3<-c("oligo","meso/eu","meso/eu","hyper")
ts_2<-c("oligo/meso","oligo/meso","eu/hyper","eu/hyper")
co<-c("<= 0.2",">2-7",">7-30",">30")
xdf<-data.frame(ts_4,ts_3,ts_2,co)
names(xdf)<-c("Trophic State (4)","Trophic State (3)","Trophic State (2)","Cut-off")
kable(xdf)
```

@@

```{r VarSel_Model1,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 1\\label{tab:VarSel_Model1}",cache=TRUE}
kable(sumTable(all_ts4[1:100]),row.names=F)
```

```{r Confusion_Model1,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 1\\label{tab:Confusion_Model1}",cache=TRUE}
kable(formatC(round(all_ts4_rf$confusion,2)),row.names=F)
```

@@

```{r Importance_Model1,results="asis",echo=FALSE, fig.cap="Importance plot for Model 1\\label{fig:Importance_Model1}",cache=TRUE}
all_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts4_vars]
importancePlot(all_ts4_rf,'gini',size=10,aes(colour=all_ts4_color))
```

@@

```{r VarSel_Model2,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 2",cache=TRUE}
kable(sumTable(all_ts3[1:100]),row.names=F)
```

```{r Confusion_Model2,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 2",cache=TRUE}
kable(formatC(round(all_ts3_rf$confusion,2)),row.names=F)
```

```{r Importance_Model2,results="asis",echo=FALSE, fig.cap="Importance plot for Model 2",cache=TRUE}
all_ts3_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts3_vars]
importancePlot(all_ts3_rf,'gini',size=10,aes(colour=all_ts3_color))
```

@@

```{r all_TS2_rf,echo=FALSE,eval=FALSE,cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts2_vars<-unique(unlist(all_ts2[1:100]))
all_ts2_rf<-runRandomForest(hkm2014Data[all_ts2_vars],hkm2014Data$TS_CHLA_2, ntree=10000,importance=TRUE,proximity=TRUE)
```

@@

```{r VarSel_Model3,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 3",cache=TRUE}
kable(sumTable(all_ts2[1:100]),row.names=F)
```

```{r Confusion_Model3,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 3",cache=TRUE}
kable(formatC(round(all_ts2_rf$confusion,2)),row.names=F)
```

```{r Importance_Model3,results="asis",echo=FALSE, fig.cap="Importance plot for Model 3",cache=TRUE}
all_ts2_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts2_vars]
importancePlot(all_ts2_rf,'gini',size=10,aes(colour=all_ts2_color))
```

@@

```{r VarSel_Model4,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 4",cache=TRUE }
kable(sumTable(gis_ts4[1:100]),row.names=F)
```

```{r Confusion_Model4,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 4",cache=TRUE }
kable(formatC(round(gis_ts4_rf$confusion,2)),row.names=F)
```

```{r Importance_Model4,results="asis",echo=FALSE, tab.cap="Importance plot for Model 4",cache=TRUE}
gis_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts4_vars]
importancePlot(gis_ts4_rf,'gini',size=10,aes(colour=gis_ts4_color))
```

@@

```{r VarSel_Model5,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 5",cache=TRUE}
kable(sumTable(gis_ts3[1:100]),row.names=F)
```

```{r Conufsion_Model5,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 5",cache=TRUE}
kable(formatC(round(gis_ts3_rf$confusion,2)),row.names=F)
```

```{r Importance_Model5,results="asis",echo=FALSE, fig.cap="Importance plot for Model 5",cache=TRUE}
gis_ts3_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts3_vars]
importancePlot(gis_ts3_rf,'gini',size=10,aes(colour=gis_ts3_color))
```

@@

```{r VarSel_Model6,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 6",cache=TRUE}
kable(sumTable(gis_ts2[1:100]),row.names=F)
```

```{r Confusion_Model6,results="asis",echo=FALSE, tab.cap="Random forest confusion matrix for Model 6",cache=TRUE}
kable(formatC(round(gis_ts2_rf$confusion,2)),row.names=F)
```

```{r Importance_Model6,results="asis",echo=FALSE, fig.cap="Importance plot for Model 6",cache=TRUE}
gis_ts2_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts2_vars]
importancePlot(gis_ts2_rf,'gini',size=10,aes(colour=gis_ts2_color))
```

@@

```{r ts_4_biov, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 4 trophic state classes",cache=TRUE}
plotCdf(hkm2014Data$TS_CHLA_4,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:9],
        y='Percent',x='Log10(Cyanobacteria Biovolume)',
        title=expression(paste('CDF for Chlorophyll ', 
                                italic("a"),' Trophic States (4 Categories)')),
        color="Trophic State\nCategories")
```
\newpage
```{r ts_3_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 3 trophic state classes",cache=TRUE }
plotCdf(hkm2014Data$TS_CHLA_3,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:8],
       y='Percent',x='Log10(Cyanobacteria Biovolume)', 
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (3 Categories)')),
       color="Trophic State\nCategories")
```
\newpage
```{r ts_2_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 2 trophic state classes",cache=TRUE}
plotCdf(hkm2014Data$TS_CHLA_2,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:7],
       y='Percent',x='Log10(Cyanobacteria Biovolume)',
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (2 Categories)')),
       color="Trophic State\nCategories")
```
\newpage
```{r scatterplot,echo=FALSE, fig.width=7, fig.cap="Cholorphyll *a* and cyanobacteria biovolume scatterplot",cache=TRUE}
scp_df<-data.frame(chla=hkm2014Data[["CHLA"]],biovp1=hkm2014Data[["sumBioV"]]+1)
scatterPlot(scp_df,xvar="chla",yvar="biovp1",zissou[8],zissou[7],zissou[6],
             title=expression(paste("Chlorophyll ", 
                                    italic("a")," and Cyanobacteria Relationship")),
             x=expression(paste('Log10(Chl ', italic("a"),')')),
             y="Log10(Cyanobaterial Biovolumes + 1)")
```
